### 2024-07-01 causal-inference.qmd
20240701 - mediation DAG
20240701 - dag modifications
20240701 - causal diagrams and formatting

As described in @sec-correlationCausation, there are several reasons why two variables, $X$ and $Y$, might be correlated:

- $X$ causes $Y$
- $Y$ causes $X$
- a third variable (i.e., confound), $Z$, influences both $X$ and $Y$
- the association between $X$ and $Y$ is spurious

@@

The third criterion involves ruling out alternative reasons why the purported cause and effect may be related.
As noted in @sec-correlationCausality, there are four reasons why $X$ may be correlated with $Y$.
If we meet the first criterion of causality, we have removed the possibility that $Y$ causes $X$ (i.e., reverse causality).
To meet the third criterion of causality, we need to remove the possibility that the association reflects a third variable (confound) that influences both the cause and effect, and we need to remove the possibility that the association is spurious—the possibility that the association between the purported cause and effect is due to random chance.
There are multiple approaches to meeting the third criterion of causality, such as by use of [experiments](#sec-causalInferenceExperiment), [longitudinal designs](#sec-causalInferenceLongitudinal), [control variables](#sec-causalInferenceControlVariables), [within-subject designs](#sec-causalInferenceWithinSubject), and [genetically informed designs](#sec-causalInferenceGeneticallyInformed), as described in @sec-approachesCausalInference.

@@

A key tool when describing a research question or hypothesis is to create a conceptual depiction of the hypothesized causal processes.
A causal diagram depicts the hypothesized causal processes that link two or more variables.
A common form of causal diagrams is the directed acyclic graph (DAG).
DAGs provide a helpful tool to communicate about causal questions and help identify how to avoid bias (i.e., over-estimation) in associations between variables due to confounding (i.e., common causes) [@Digitale2022].
You can use the `R` package `dagitty` [@Textor2017] or the associated browser-based extension, DAGitty: https://dagitty.net (archived at https://perma.cc/U9BY-VZE2) to create DAGs.
Examples of various causal diagrams that could explain why *X* is associated with *Y* are in @fig-XCausesY, @fig-YCausesX, and @fig-ZCausesXandY.

```{r}
#| label: fig-XCausesY
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting *X* Causing *Y*."

@@

```{r}
#| eval: false

@@

plot(XCausesY_alt)
ggdag(XCausesY_alt)
```

```{r}
#| label: fig-YCausesX
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting Reverse Causation: *Y* Causing *X*."

@@

```{r}
#| eval: false

@@

plot(YCausesX_alt)
ggdag(YCausesX_alt)
```

```{r}
#| label: fig-ZCausesXandY
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting a Third Variable Confound, *Z*, Causing *X* and *Y*, Thus Explaining Why *X* and *Y* are associated."

@@

```{r}
#| eval: false

@@

plot(ZCausesXandY_alt)
ggdag(ZCausesXandY_alt)
```

@@

mediationDag <- dagitty::dagitty("dag{
  X -> M
  X -> A
  M -> Y
  Y -> Z
  A <-> M
}")

@@

dagitty::adjustmentSets(
  mediationDag, 
  exposure = "X",
  outcome = "Z",
  effect = "total")
```

```{r}
#| eval: false

mediationDag_alt <- ggdag::dagify(
  M ~ X,
  Y ~ M,
  Z ~ Y,
  A ~ X,
  A ~~ M
)

plot(mediationDag_alt)
ggdag(mediationDag_alt)
```

@@

For instance, we might observe that sports drink consumptions is concurrently associated with better player performance.
Among [observational/correlational research designs](#sec-correlationalStudy), [cross-sectional designs](#sec-crossSectional) tend to have the weakest [internal validity](#sec-internalValidity).
For the reasons described in @sec-correlationCausality, if we observe a cross-sectional association between $X$ (e.g., sports drink consumption) and $Y$ (e.g., player performance), we have little confidence that $X$ causes $Y$.
As a result, [longitudinal designs](#sec-longitudinal) can be valuable for more closely approximating causality if an [experimental designs](#sec-causalInferenceExperiment) is not possible.
Consider a lagged association that might be observed in a [longitudinal design](#sec-longitudinal), as in @fig-laggedAssociation, which is a slightly better approach than relying on cross-sectional associations:

@@

One of the plausible alternatives to the inference that $X$ causes $Y$ is that there are third variable confounds that influence both $X$ and $Y$, thus explaining why $X$ and $Y$ are associated, as depicted in @fig-correlationAndCausation3 and @fig-ZCausesXandY.
Thus, another approach that can help increase [internal validity](#sec-internalValidity) is to include plausible confounds as control variables.
For instance, if a third variable such as education level might be a confound that influences both sports drink consumption and player performance, you could include education level as a [covariate](#sec-covariates) in the model.
Inclusion of a [covariate](#sec-covariates) attempts to control for the variable by examining the association between the [predictor variable](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy) while holding the [covariate](#sec-covariates) variables constant.
For instance, such a model would examine whether, when accounting for education level, there is an association between sports drink consumption and player performance.

>>>>

As described in @sec-correlationCausation, there are several reasons why two variables, `X` and `Y`, might be correlated:

- `X` causes `Y`
- `Y` causes `X`
- a third variable (i.e., confound), `Z`, influences both `X` and `Y`
- the association between `X` and `Y` is spurious

@@

The third criterion involves ruling out alternative reasons why the purported cause and effect may be related.
As noted in @sec-correlationCausality, there are four reasons why `X` may be correlated with `Y`.
If we meet the first criterion of causality, we have removed the possibility that `Y` causes `X` (i.e., reverse causality).
To meet the third criterion of causality, we need to remove the possibility that the association reflects a third variable (confound) that influences both the cause and effect, and we need to remove the possibility that the association is spurious—the possibility that the association between the purported cause and effect is due to random chance.
There are multiple approaches to meeting the third criterion of causality, such as by use of [experiments](#sec-causalInferenceExperiment), [longitudinal designs](#sec-causalInferenceLongitudinal), [control variables](#sec-causalInferenceControlVariables), [within-subject designs](#sec-causalInferenceWithinSubject), and [genetically informed designs](#sec-causalInferenceGeneticallyInformed), as described in @sec-approachesCausalInference.

@@

A key tool when describing a research question or hypothesis is to create a conceptual depiction of the hypothesized causal processes.
A causal diagram depicts the hypothesized causal processes that link two or more variables.
A common form of causal diagrams is the directed acyclic graph (DAG).
DAGs provide a helpful tool to communicate about causal questions and help identify how to avoid bias (i.e., over-estimation) in associations between variables due to confounding (i.e., common causes) [@Digitale2022].
You can use the `R` package `dagitty` [@Textor2017] or the associated browser-based extension, DAGitty: https://dagitty.net (archived at https://perma.cc/U9BY-VZE2) to create DAGs.
Examples of various causal diagrams that could explain why `X` is associated with `Y` are in @fig-XCausesY, @fig-YCausesX, and @fig-ZCausesXandY.

```{r}
#| label: fig-XCausesY
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting `X` Causing `Y`."

@@

Here is an alternative way of specifying the same diagram (more similar to `lavaan` syntax):

```{r}
#| label: fig-XCausesY2
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting `X` Causing `Y`."

@@

#plot(XCausesY_alt) # this creates the same plot as above
ggdag::ggdag(XCausesY_alt) + theme_dag_blank()
```

```{r}
#| label: fig-YCausesX
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting Reverse Causation: `Y` Causing `X`."

@@

Here is an alternative way of specifying the same diagram (more similar to `lavaan` syntax):

```{r}
#| label: fig-YCausesX2
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting Reverse Causation: `Y` Causing `X`."

@@

#plot(YCausesX_alt) # this creates the same plot as above
ggdag::ggdag(YCausesX_alt)
```

```{r}
#| label: fig-ZCausesXandY
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting a Third Variable Confound, `Z`, Causing `X` and `Y`, Thus Explaining Why `X` and `Y` are associated."

@@

Here is an alternative way of specifying the same diagram (more similar to `lavaan` syntax):

```{r}
#| label: fig-ZCausesXandY2
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting a Third Variable Confound, `Z`, Causing `X` and `Y`, Thus Explaining Why `X` and `Y` are associated."

@@

#plot(ZCausesXandY_alt) # this creates the same plot as above
ggdag::ggdag(ZCausesXandY_alt)
```

@@

mediationDag <- dagitty::dagitty("dag{
  X -> M1
  X -> M2
  M1 -> Y
  M2 -> Y
  M1 <-> M2
}")

@@

dagitty::adjustmentSets(
  mediationDag, 
  exposure = "M1",
  outcome = "Y",
  effect = "total")
```

In this example, `X` influences `Y` via `M1` and `M2` (i.e., multiple mediatirs), and `M1` is also associated with `M2`.
The `dagitty::impliedConditionalIndependencies()` function identifies variables in the causal diagram that are conditionally independent (i.e., uncorrelated) after controlling for other variables in the model.
For this causal diagram, `X` is conditionally independent with `Y` because `X` is independent with `Y` when controlling for `M1` and `M2`.

The `dagitty::adjustmentSets()` function identifies variables that would be necessary to control for (i.e., to include as covariates) in order to identify an unbiased estimate of the association (whether the total effect, i.e., `effect = "total"`; or the direct effect, i.e., `effect = "direct"`) between two variables (`exposure` and `outcome`).
In this case, to identify the unbiased association between `M1` and `Y`, it is important to control for `M2.`

Here is an alternative way of specifying the same diagram (more similar to `lavaan` syntax):

```{r}
#| label: fig-dag2
#| fig-cap: "Causal Diagram (Directed Acyclic Graph)."

mediationDag_alt <- ggdag::dagify(
  M1 ~ X,
  M2 ~ X,
  Y ~ M1,
  Y ~ M2,
  M1 ~~ M2
)

#plot(mediationDag_alt) # this creates the same plot as above
ggdag::ggdag(mediationDag_alt)
```

@@

For instance, we might observe that sports drink consumptions is concurrently associated with better player performance.
Among [observational/correlational research designs](#sec-correlationalStudy), [cross-sectional designs](#sec-crossSectional) tend to have the weakest [internal validity](#sec-internalValidity).
For the reasons described in @sec-correlationCausality, if we observe a cross-sectional association between `X` (e.g., sports drink consumption) and `Y` (e.g., player performance), we have little confidence that `X` causes `Y`.
As a result, [longitudinal designs](#sec-longitudinal) can be valuable for more closely approximating causality if an [experimental designs](#sec-causalInferenceExperiment) is not possible.
Consider a lagged association that might be observed in a [longitudinal design](#sec-longitudinal), as in @fig-laggedAssociation, which is a slightly better approach than relying on cross-sectional associations:

@@

One of the plausible alternatives to the inference that `X` causes `Y` is that there are third variable confounds that influence both `X` and `Y`, thus explaining why `X` and `Y` are associated, as depicted in @fig-correlationAndCausation3 and @fig-ZCausesXandY.
Thus, another approach that can help increase [internal validity](#sec-internalValidity) is to include plausible confounds as control variables.
For instance, if a third variable such as education level might be a confound that influences both sports drink consumption and player performance, you could include education level as a [covariate](#sec-covariates) in the model.
Inclusion of a [covariate](#sec-covariates) attempts to control for the variable by examining the association between the [predictor variable](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy) while holding the [covariate](#sec-covariates) variables constant.
For instance, such a model would examine whether, when accounting for education level, there is an association between sports drink consumption and player performance.

### 2024-07-01 getting-started.qmd
20240627 - sessionInfo()
20240629 - fix error
20240701 - causal diagrams and formatting

A full outer join includes all rows in $x$ **or** $y$.
It returns columns from $x$ and $y$.
Here is how to merge two data frames using a full outer join (i.e., "full join"):

@@

A left outer join includes all rows in $x$.
It returns columns from $x$ and $y$.
Here is how to merge two data frames using a left outer join ("left join"):

@@

A right outer join includes all rows in $y$.
It returns columns from $x$ and $y$.
Here is how to merge two data frames using a right outer join ("right join"):

@@

An inner join includes all rows that are in **both** $x$ **and** $y$.
An inner join will return one row of $x$ for each matching row of $y$, and can duplicate values of records on either side (left or right) if $x$ and $y$ have more than one matching record.
It returns columns from $x$ and $y$.
Here is how to merge two data frames using an inner join:

@@

A semi join is a filter.
A left semi join returns all rows from $x$ **with** a match in $y$.
That is, it filters out records from $x$ that are not in $y$.
Unlike an [inner join](#innerJoin), a left semi join will never duplicate rows of $x$, and it includes columns from only $x$ (not from $y$). 
Here is how to merge two data frames using a left semi join:

@@

An anti join is a filter.
A left anti join returns all rows from $x$ **without** a match in $y$.
That is, it filters out records from $x$ that are in $y$.
It returns columns from only $x$ (not from $y$).
Here is how to merge two data frames using a left anti join:

@@

A cross join combines each row in $x$ with each row in $y$.

@@

```{r}
sessionInfo
```

>>>>

A full outer join includes all rows in `x` **or** `y`.
It returns columns from `x` and `y`.
Here is how to merge two data frames using a full outer join (i.e., "full join"):

@@

A left outer join includes all rows in `x`.
It returns columns from `x` and `y`.
Here is how to merge two data frames using a left outer join ("left join"):

@@

A right outer join includes all rows in `y`.
It returns columns from `x` and `y`.
Here is how to merge two data frames using a right outer join ("right join"):

@@

An inner join includes all rows that are in **both** `x` **and** `y`.
An inner join will return one row of `x` for each matching row of `y`, and can duplicate values of records on either side (left or right) if `x` and `y` have more than one matching record.
It returns columns from `x` and `y`.
Here is how to merge two data frames using an inner join:

@@

A semi join is a filter.
A left semi join returns all rows from `x` **with** a match in `y`.
That is, it filters out records from `x` that are not in `y`.
Unlike an [inner join](#innerJoin), a left semi join will never duplicate rows of `x`, and it includes columns from only `x` (not from `y`). 
Here is how to merge two data frames using a left semi join:

@@

An anti join is a filter.
A left anti join returns all rows from `x` **without** a match in `y`.
That is, it filters out records from `x` that are in `y`.
It returns columns from only `x` (not from `y`).
Here is how to merge two data frames using a left anti join:

@@

A cross join combines each row in `x` with each row in `y`.

@@

```{r}
sessionInfo()
```

### 2024-07-01 research-methods.qmd
20240627 - sessionInfo() for reproducibility
20240627 - sessionInfo()
20240701 - causal diagrams and formatting

Just because $X$ is associated with $Y$ does not mean that $X$ causes $Y$.
Consider that you find an association between variables $X$ and $Y$.

There are several reasons why you might observe an association between $X$ and $Y$:

- $X$ causes $Y$
- $Y$ causes $X$
- a third variable (i.e., confound), $Z$, influences both $X$ and $Y$
- the association between $X$ and $Y$ is spurious

For instance, one possibility is that the association we observed reflects our hypothesis that $X$ causes $Y$, as depicted in @fig-correlationAndCausation1.
That is, consumption of more sports drink may improve players' performance.

@@

Hypothesized Causal Effect Based on an Observed Association Between $X$ and $Y$, Such That $X$ Causes $Y$.
:::

However, a second possibility is that the association reflects the opposite direction of effect, where $Y$ actually causes $X$, as depicted in @fig-correlationAndCausation2.
For instance, greater performance may lead players to drink more sports drink (rather than the reverse).

@@

Reverse (Opposite) Direction of Effect From the Hypothesized Effect, Where $Y$ Causes $X$.
:::

A third possibility is that the association could reflect the influence of a third variable.
If a third variable is a common cause of each and accounts for their association, it is a *confound*.
An observed association between $X$ and $Y$ could reflect a confound—i.e., a cause ($Z$) that influences both $X$ and $Y$, which explains why $X$ and $Y$ are correlated even though they are not causally related.
A third variable confound that is a common cause of both $X$ and $Y$ is depicted in @fig-correlationAndCausation3.
For instance, it may not be that sport drink consumption per se influences player performance; rather, it may be that players who are more intelligent or have more financial resources tend to drink more sports drinks and also tend to perform better.
In this case, intelligence or financial resources may be a confound that influences both sports drink consumption and player performance, but sports drink consumptions—though correlated with player performance—does not influence player performance.

@@

Confounded Association Between $X$ and $Y$ due to a Common Cause, $Z$.
:::

@@

>>>>

Just because `X` is associated with `Y` does not mean that `X` causes `Y`.
Consider that you find an association between variables `X` and `Y`.

There are several reasons why you might observe an association between `X` and `Y`:

- `X` causes `Y`
- `Y` causes `X`
- a third variable (i.e., confound), `Z`, influences both `X` and `Y`
- the association between `X` and `Y` is spurious

For instance, one possibility is that the association we observed reflects our hypothesis that `X` causes `Y`, as depicted in @fig-correlationAndCausation1.
That is, consumption of more sports drink may improve players' performance.

@@

Hypothesized Causal Effect Based on an Observed Association Between `X` and `Y`, Such That `X` Causes `Y`.
:::

However, a second possibility is that the association reflects the opposite direction of effect, where `Y` actually causes `X`, as depicted in @fig-correlationAndCausation2.
For instance, greater performance may lead players to drink more sports drink (rather than the reverse).

@@

Reverse (Opposite) Direction of Effect From the Hypothesized Effect, Where `Y` Causes `X`.
:::

A third possibility is that the association could reflect the influence of a third variable.
If a third variable is a common cause of each and accounts for their association, it is a *confound*.
An observed association between `X` and `Y` could reflect a confound—i.e., a cause (`Z`) that influences both `X` and `Y`, which explains why `X` and `Y` are correlated even though they are not causally related.
A third variable confound that is a common cause of both `X` and `Y` is depicted in @fig-correlationAndCausation3.
For instance, it may not be that sport drink consumption per se influences player performance; rather, it may be that players who are more intelligent or have more financial resources tend to drink more sports drinks and also tend to perform better.
In this case, intelligence or financial resources may be a confound that influences both sports drink consumption and player performance, but sports drink consumptions—though correlated with player performance—does not influence player performance.

@@

Confounded Association Between `X` and `Y` due to a Common Cause, `Z`.
:::

@@

::: {.content-visible when-format="html"}

## Session Info {#sec-researchMethodsSessionInfo}

```{r}
sessionInfo()
```

:::
